SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2020-01-17 2017-Xudong-SMT/ActaInformatica.tex
Add Alg.3 and Alg.4 according to the 2nd comment of Reviewer #1.

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=\textit{filterUnsatTransitions}($\mathit{L,Assignments}$);
\State $\mathit{OA_1^0}$=\textit{makeReachableSubAutomaton}($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_1^n}$=\textit{RefineReachableSubAutomaton}($\mathit{OA_1^{n-1}}$)
\Until  $\mathit{OA_1^n}=\mathit{OA_1^{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_1^n}$;

@@

\begin{algorithm}[h]
\caption{\emph{MakeTransitions}() for a pNet node}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $P$ with subnets $\overline{sn}$ and holes $\overline{hole}$; a global state $S$.
\State Initialize empty list $l$ and set $L$ for sub-transitions and transitions, respectively;
	\For{each $\mathit{Subnet}$ in $\overline{sn}$}
        \hfill $\setminus \setminus$ Recursively apply the semantic rules on the subnets
           \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\State $\overline{comb} = \mathit{Combining}(l)$;
	\For{ each $sv \in SV$ and each $comb \in \overline{comb}$ }
           \State $ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \overline{\mathit{hole}})$;
           \IfThen{($ot$ is defined)}{Store $ot$ in $L$; ~~~ $\setminus \setminus$ if $\mathit{Matching}()$ succeeds}
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{RefineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
variable $v$.

@@

\QIN{As the removal of the unsatisfiable transitions \YX{may entail some transitions without preceeding} transitions, }
the set $L1$ might contain unreachable transitions. To remove those transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$.
%as defined in page~\pageref{def-states}
More precisely, this is a residual algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

@@

\subsection{Computing Reachable State Space on the Fly}

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=\textit{filterUnsatTransitions}($\mathit{L,Assignments}$);
\State $\mathit{OA_2^0}$=\textit{makeReachableSubAutomaton}($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_2^n}$=\textit{RefineReachableSubAutomaton}($\mathit{OA_2^{n-1}}$)
\Until  $\mathit{OA_2^n}=\mathit{OA_2^{n-1}}$;
\State \Return $\mathit{OA_2^n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

@@

The easy case is when the system does not contain assignments to state
variables. This does not mean ``data-independency'': we can still have
guards in the PLTS transitions and in the synchronisation vectors,
making the behaviour depending on the values of input variables. But
the open automaton needs not to collect the set of possible values dealing
with the values of local variables, so the final automaton is $AO_2^0$ as
computed in line 13.

>>>>

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a PLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$;
          add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State $\mathit{L1}$=\textit{filterUnsatTransitions}($\mathit{L,Assignments}$);
\State $\mathit{OA_1^0}$=\textit{makeReachableSubAutomaton}($\mathit{InitState}(P),L1)$;
\Repeat $\mathit{OA_1^n}$=\textit{refineReachableSubAutomaton}($\mathit{OA_1^{n-1}}$)
\Until  $\mathit{OA_1^n}=\mathit{OA_1^{n-1}}$;   \hfill $\setminus \setminus$ Fixed point reached
\State \Return $\mathit{OA_1^n}$;

@@

\begin{algorithm}[h]
\caption{\emph{MakeTransitions}() for a pNet node}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $P$ with subnets $\overline{sn}$ and holes $\overline{hole}$; a global state $S$.
\State Initialize an empty list $l$ and a set $L$ for sub-transitions and transitions, respectively;
	\For{each $\mathit{Subnet}$ in $\overline{sn}$}
        \hfill $\setminus \setminus$ Recursively apply the semantic rules on the subnets
           \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\State $\overline{comb} = \mathit{Combining}(l)$;
	\For{ each $sv \in SV$ and each $comb \in \overline{comb}$ }
           \State $ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \overline{\mathit{hole}})$;
           \IfThen{($ot$ is defined)}{Store $ot$ in $L$; ~~~ $\setminus \setminus$ if $\mathit{Matching}()$ succeeds}
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{refineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
variable $v$.

@@

\QIN{As the removal of the unsatisfiable transitions \YX{may entail some transitions without predecessor} transitions, }
the set $L1$ might contain unreachable transitions. To remove those transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$.
%as defined in page~\pageref{def-states}
More precisely, this is a residual algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

@@

\begin{algorithm}[htbp]
  \caption{\QINX{makeReachableSubAutomaton()}}
  \label{alg:makeReachableSubAutomaton}
\begin{algorithmic}[1]
\Require \QINX{The initial state of the pNet $\textit{InitState}(P)$. A set of satisfiable transitions $L$.}
\State Initialize a set $A$;
\State Store $(\textit{InitState}(P), \textit{InitState}(P).Assignments)$ into $S$;  \hfill $\setminus \setminus$$S.Assignments$ is the assignment of the variable in the state $S$.
\While{!$\mathit{isEmpty}(A)$}
	\For {each $(S, Assignments) \in A$}
          \For {each $OT$ from $S$}
          \State Check the predicate of $OT$ taking $Assignments$ in to account;
	  \IfThenElse{$\textit{false}$}{\textbf{skip};}{Add target of $OT$ and its assignments ($S'$, $S'.Assignments$) into $A$;}
	\EndFor	\EndFor
\EndWhile
\State Build $\mathit{OA}$ with the set $L$;
\State \QINX{\Return $\mathit{OA}$;}
\end{algorithmic}  
\end{algorithm}

\begin{algorithm}[htbp]
  \caption{\QINX{refineReachableSubAutomaton()}}
  \label{alg:refineReachableSubAutomaton}
\begin{algorithmic}[1]
\Require \QINX{An open automaton $\mathit{OA}$.}
\Repeat
\For {\QINX{each global state $S$ in $\mathit{OA}$}}
          \State \QINX{$b=$ whether $S$ has preceeding transitions.}
	  \IfThenElse{$b$}{\textbf{skip};}{\QINX{remove $S$ and the transitions from $S$;}}
\EndFor
\Until  \QINX{$\mathit{OA}$ does not change;}
\State \QINX{\Return $\mathit{OA}$;}
\end{algorithmic}  
\end{algorithm}

\subsection{Computing Reachable State Space on-the-fly}

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=\textit{filterUnsatTransitions}($\mathit{L,Assignments}$);
\State $\mathit{OA_2^0}$=\textit{makeReachableSubAutomaton}($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_2^n}$=\textit{refineReachableSubAutomaton}($\mathit{OA_2^{n-1}}$)
\Until  $\mathit{OA_2^n}=\mathit{OA_2^{n-1}}$;
\State \Return $\mathit{OA_2^n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

@@

The easy case is when the system does not contain assignments to state
variables. This does not mean ``data-independency'': we can still have
guards in the PLTS transitions and in the synchronisation vectors,
making the behaviour depending on the values of input variables. But
the open automaton needs not to collect the set of possible values dealing
with the values of local variables, so the final automaton is $OA_2^0$ as
computed in line 13.
